---
title: "R code and report"
author: "BXXXXXX"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## --- FIRST THING ---

> Delete the task + report (Task 1 or Task 2) that you are **not** completing in R and then delete this FIRST THING section. 

# Task 1: R 

In this task you are asked to investigate the data in `scot_unintentional_injuries.csv` and `healthboards.csv`. 

The Unintentional Injuries dataset from Public Health Scotland provides information on emergency hospital admissions as a result of unintentional injuries and assaults. You can find more [information about the unintentional injuries data set here, including a data dictionary](https://www.opendata.nhs.scot/dataset/unintentional-injuries/resource/aee43295-2a13-48f6-bf05-92769ca7c6cf) and [here for more information about the health boards dataset](https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc). Do not import the data via the URL, you must use the data files provided for this assignment. 

Follow the steps below to ultimately answer the final questions with the data. 

* load the necessary libraries and packages
* read in the data
* join the two data sets together
* check the data types of the variables of interest and ensure they are they type you want them to be. If not, change them.
* check for the content/information contained within your variables of interest and ensure you are satisfied
with the presentation
* check for and deal with any missing data, if deemed appropriate
* check the data formats (wide vs long) and transform if it is not appropriate

*Question:* **Among people aged 65 and older in 2024/25, did NHS Lothian and NHS Greater Glasgow and Clyde have the same 4 most common causes of accidental injuries leading to emergency hospital admissions? What were they (the 4 most common causes of accidental injury in these regions)?**

```{r}

```

```{r}

```

```{r}

```

**Task 1 (R) Answer** (in 1-2 sentences): 


## Task 1 Report

* word limit: 1100 words maximum including Python report  
* word count: **add here once you have completed this section**



# Task 2: R

There has been an IT failure in the prescribing databases in a local area. The only remaining data around medications is the backup file below, but it has been scrambled.

Explore the following data object `medications`. We know from notes that the data should include information on pill shape, color, where it was prescribed, when it was prescribed, the BNF (British National Formulary) code, dose frequency, dose, unit of measure of dose, maximum safe temperature for storage in Celsius, and if it needs to be refrigerated.

Select 7 of the 10 elements of data and reconstruct the dataset about medications. Present `medications` in a clearer and better structured format, ensuring that the data structure and data types are appropriate, given the (limited) information provided.

The code to load the data object `medications` has been provided below.

```{r task2-data-r}
# install.packages("rjson")
# uncomment above if you have not installed the package before 

library(rjson)

medications <- fromJSON(file = "data/task2_data.json")
```

```{r}

```

```{r}

```


# Task 2 Report

* word limit: 1100 words maximum including Python report  
* word count: **add here once you have completed this section**


