---
title: "Error of the week: Pivoting in R"
author: "Dr Brittany Blankinship"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages}
library(tidyverse)
```

## Pivoting data frames 

Lets create an toy data set representing a questionnaire gathered from 20 participants. 5 questions were asked alongside demographic information of age and sex. 

```{r create data}
data_df <- tibble(id = 1:20,
                  age = sample(20:80, 20, replace = TRUE),
                  sex = sample(c("M", "F", "Not specified"), 20, 
                               prob = c(0.48, 0.49, 0.03), replace = TRUE),
                  Q_1 = round(runif(n = 20, min = 1, max = 12)),
                  Q_2 = round(runif(n = 20, min = 1, max = 15)),
                  Q_3 = round(runif(n = 20, min = 1, max = 7)),
                  Q_4 = round(runif(n = 20, min = 1, max = 12)),
                  Q_5 = round(runif(n = 20, min = 1, max = 13)))

data_df
```

First thing is I want to change my data types and make `id` and `sex` a factor:

``` {r manipulate}
data_df <- data_df |>
  mutate(id = as.factor(id), # need to use as.factor here as fct will not take an integer vector, only a character vector 
         sex = fct(sex))

data_df

# change on line 39 as.factor() to fct() and see if you can understand the error message 
```


Now I want to pivot the dataframe to be long with a column called `Question` for the questions asked (1-5) and then a column `Answer` with the respondent's answers:

``` {r pivot-long-error1}

data_long <- data_df |>
  pivot_longer(names_to = "Question", 
               values_to = "Answer")
```

Ah an error! "Error in `pivot_longer()`: ! `cols` must select at least one column." 

This suggests that the `cols` argument is empty. Let's pull up the documentation of pivot_longer 

```{r pl-docs}
?pivot_longer
```

We can see that `cols` is "<tidy-select> Columns to pivot into longer format.", but we have not specified any columns in this argument in line 53! In the documentation it also highlights that we can use tidyselect helpers, so let's try that.

*📝 Note* on Noteable, there is a known-bug that when you click a link in a help file, there is an error. To see the documentation around tidy-select helpers use this:

```{r tidy-select-docs}
?tidyr_tidy_select
```


``` {r pivot-long-everything}
data_long <- data_df |>
  pivot_longer(cols = everything(), 
               names_to = "Question", 
               values_to = "Answer")

```

Ah another error! But this time a different one. Here we are told that "Error in `pivot_longer()`: ! Can't combine `id` <factor<3be6f>> and `age` <integer>." 
So there issue is one of data types! As we are wanting all values to be in the `Answer` column, the pivoted columns need to be of the same data type.

At the moment, `age` is an integer and `id` a factor. We know that our questions columns are numeric as well, so what if we say pivot columns that are **not** `id` instead... to do this we can use `!` as the not operator.   

``` {r pivot-long-error3}
data_long <- data_df |>
  pivot_longer(cols = !id, 
               names_to = "Question", 
               values_to = "Answer")

```
Another error! This time the same error message as above, but with a different set of columns causing the problem. "Error in `pivot_longer()`: ! Can't combine `age` <integer> and `sex` <factor<70d94>>."

We also have the `sex` column, which is a factor. So let's add that to our columns **not** to include in pivoting. To do this, we can use `c()` to create a vector and pass this to the `cols` argument. 

``` {r pivot-long}
data_long <- data_df |>
  pivot_longer(cols = !c(id,sex), 
               names_to = "Question", 
               values_to = "Answer")

data_long
```

Woo no error! But in checking out output, this is still not correct. This is one of the reasons it is **always important to check your output** and not to rely solely on error messages to tell you if you code is not doing what you expect. As the programmer, you need to use your critical thinking skills to assess if your code functioning properly. 

The pivoting took place, however, age should not be included in the question column. Age is an ID variable, not an observation variable (i.e., variables we want to pivot). Our ID variables are `id`, `sex`, and `age`. So we need to pass a vector to the `cols` argument so that they are treated as ID variables, and not included in pivoting. 

``` {r pivot-long-solution}
data_long_correct <- data_df |>
  pivot_longer(cols = !c(id, sex, age), 
               names_to = "Question", 
               values_to = "Answer")

data_long_correct
```

We can see indeed that our data is in the long format we expect, with our ID variables being repeated 5 times, one row for each question. And the dimensions of our dataframe is as expected 100 x 5 - 100 rows and 5 columns. This is expected as 20 (number of participants) x 5 (number of questions) = 100 rows and we have 3 ID columns plus `Question` and `Answer` that we made when pivoting = 5 columns. 

## Success!! 🎉

---
*Dr. Brittany Blankinship (2026)* 

