---
title: "R data types notebook"
subtitle: "Numeric data"
author: "Dr Brittany Blankinship"
date: "`r Sys.Date()`"
toc: true
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages, echo=FALSE, message=FALSE}
# first lets load the libraries we need for this notebook 

Sys.setenv(TZ = "Europe/London") # set timezone (TZ argument) to whatever timezone you may be in. The allows the to packages to work as intended

library(tidyverse) # as the tidyverse is a whole suite of packages, we just need one lirbary call to load all that we need 
```


## 1. Introduction 

In these notebooks we will be looking at the data types discussed in the Week 2 & 3 content, how they work in R, and how you can manipulate them. A data type of an object or variable specifies the type of data that is stored inside that object and how you can then interact with it.

✅  *Reminder* that you can open the RMarkdown document outline at the top of this panel to navigate to different sections in this notebook. You can also collapse sections my clicking on the little arrow that appears by the line number. Note though that if you collapse a level 2 section, all level 3 sections contained within it will also collapse. Try this for line 16 Introduction to see what I mean

*📝 Note* there is code throughout this workbooks which will intentionally produce an error. Try to read the error messages and see if you can understand what is causing the problem and why. Reading error messages and not being frightened by them is a key skill for any programmer! Review the materials from the Week 2 tutorial on reading error messages - remember, you are the detective! 🔍 

✅ These errors mean that you will not be able to knit the document unless you comment out the code that generates them. 

✅ Remember to RUN ALL CODE CHUNKS IN ORDER AS YOU GO THROUGH THIS NOTEBOOK (if you skip some, you might see some unexpected errors)

✅ Do not forget about the keyboard shortcuts for running code in RMD documents:

* Ctrl/Cmd + Enter to run the current line/selection of code (if highlighted)

* Ctrl/Cmd + Shift + Enter to run the entire code chunk 


### 1.1 R has 5 basic types.

* logical: TRUE, FALSE  
* numeric: 1.0, 15.3
  - double: 1.0, 15.3
  - integer: -1,2  
* complex: 1+4i  
* character: "health care", "social care"

Another important data type that helps to interact with categorical variables is 

* factor: "apple", "banana", "orange"

To work with date and time data we use the `{lubridate}` package 

* date: "2026-07-01"
* time: "15:30:43"
* datetime: "2026-07-01 15:30:43 UTC"

### 1.2 R also provides many functions to examine data type features, for example:  

* `mode()` - how is the object stored in R? 
* `class()` - what kind of object is it (high-level)?  
* `typeof()` - what is the object’s data type (low-level)?  
* `length()` - how long is the data?  

If you would like to check if an object is a particular data type, there is a series of `"is.datatype()"` functions which produce a Boolean value output:

* `is.logical()` to check if the data object holds a logical value (i.e., `TRUE` or `FALSE`)
* `is.numeric()` to check if the data object is numeric
* `is.double()` to check if the data object is a double-precision floating point number 
* `is.integer()` to check if the data object is an integer
* `is.complex()` to check if the data object is complex
* `is.character()` to check if the data object is character
* `is.factor()` to check if the data object is a nominal factor
* `is.ordered()` to check if the data object is an ordinal factor 

#### 1.2.1 datetime data 

For datetime data `{lubridate}` also has a series of `"is.X()"` functions which mirror the constructors in base R above:

* `is.Date()` to check if the data is a Date type
* `is.timespan()` to check if the data object is a length of time 


#### 1.2.2 Changing datatypes (i.e., type coercion) 

To coerce or change an object to another data type, there is a series of `as.datatype()` functions 

* `as.logical()` 
* `as.numeric()`
* `as.double()`
* `as.integer()`
* `as.complex()`
* `as.character()`

* `fct()` to encode a vector as a factor. There are `is._()` and `as._()` functions for membership and coercion 
  - `as_factor()`: note the syntax difference from the other functions (using an underscore rather than a dot). This is because we will be using the factor coercion function from the `{forcats}` package rather than base R. See the Factor section for more details 
  - `as.ordered()`: defaults to lexical ordering (alphabetical order) for coerced characters

* `as_datetime()` 
* `as_date()`

#### 1.2.3 Describing dataframes 

When working with dataframes (which we will learn more about in Week 4), there are 2 similar functions that are useful:  

* `str()` to get the structure of an R object 
* `dplyr::glimpse()` to see a more detailed glimpse of the structure of a dataframe, including a preview of the data 

*📝 Note* remember the namespace syntax which uses double colon `::` means `package::function()`? So the above on line 105 means the `glimpse()` function from the `{dplyr}` package 

## 2. Numeric 

In R, numeric is both the high level data type category for numbers as well as its own data type. Notably, when you call `class()` of a numeric variable the output will be numeric, but lower level `typeof()` will be double. This is because the default behavior in R is to store numeric data as double. Complex number data types in R are a unique mode, class, and data type. As they are not commonly used, I have not included a separate section for them however (but at the top of the document on line 50 they are not subset within numeric for this reason). 

There are some built-in functions when working with numeric data that we can use:

* `round()` to round a number to the nearest integer. 
  - **📝 Note** when the number ends in .5 `round()` has some unexpected behavior as it uses the IEC 60559 standard "go to the even digit". This means that the rounding will go to the nearest even digit. This behavior is what was referenced in the Data Dialogue with Dr Dimitrios Doudesis. You can also specify the number of decimal places with the `digits` argument in the `round()` function 
* `abs()` to get the absolute value of a number 
* `sum()` used to sum a list or vector of values
* `min()` returns the minimum value  
* `max()` returns the maximum value 

```{r rounding-behavior}
# Based on what is mentioned above in line 312, what do you expect the output to be for the following round functions? Think about this before running the code line-by-line

round(4.5)

round(5.5)

round(0.5)
```


```{r numeric-double}
dd <- 56.48

class(dd)
typeof(dd)
mode(dd)
```

```{r numeric-numeric}
ii <- 80

class(ii)
typeof(ii)
mode(ii)
```

```{r numeric-complex}
cc <- 7i + 3 # note here I put the imaginary element before the real element

cc # when printed, R automatically re-formats the number to (real + imaginary)

typeof(cc)
mode(cc)
```

You can also create numeric data with scientific notation in R but inputting `e` or `E` followed by a positive or negative integer. 

```{r numeric-scientific-1}
sci_num1 <- 4.2e-3

sci_num1

typeof(sci_num1)
```

```{r numeric-scientific-2}
sci_num2 <- .4E5

sci_num2

typeof(sci_num2)
```

### 2.1 Converting to and from numeric 

Notably, because integers and doubles are stored differently and have differing levels of precision, using the `as.integer()` function truncates floating-point numbers so that the result is the nearby integer closer to zero (rounding down in other words). 

```{r numeric-coercion-1}
# create a double 
dbl1 <- 4.8573

# check the data type 
typeof(dbl1)

# check the data class 
class(dbl1)

# check the storage mode - for numeric data this will always be numeric as the higher order class 
mode(dbl1)
```


```{r numeric-coercion-2}
#convert object dbl1 to an integer 
dbl1_int <- as.integer(dbl1) #this will truncate the value of a double 

dbl1_int

typeof(dbl1_int)
class(dbl1_int)
mode(dbl1_int)

is.numeric(dbl1_int)
is.integer(dbl1_int)
is.double(dbl1_int)
```

On a technical note, `is.numeric()` tests the mode, not the class while `as.numeric()` (which is identical to `as.double()`) coerces to the class. 

By default, numeric data types are doubles, even if you input a full number, as seen in line 338 with object `ii`. There are, however, a few ways of automatically creating integer data types in R. If you would like to create an integer variable without specifying `as.integer()` you can put an `L` after the number. You can also create an integer using `:` which tells R to generate a sequence of numbers from the value on the left to the value on the right, going up by 1 each time. 

```{r numeric-coercion-3}
ii2 <- 45L
ii2

class(ii2)
typeof(ii2)
mode(ii2) # this is numeric as numeric is the mode of storage of number based data in R for numeric, double, and integer types 
```

```{r numeric-coercion-vector}
ii_vec <- c(44L, 687L, 2L, 60L)
ii_vec

class(ii_vec)
typeof(ii_vec)
mode(ii_vec)
```

```{r numeric-coercion-4}
ii_vec2 <- c(1:15)
ii_vec2

class(ii_vec2)
typeof(ii_vec2)
mode(ii_vec2)
```

If you randomly sample a certain number of integer values within a specified range using `sample()`, this will be automatically typed as an integer as well. If you input a double (or number with a decimal place) as the first argument in `sample()` this will, as expected, produce a vector of type double. 

Run code the code chunks on lines 442 (numeric-coercion-5) and 451 (numeric-coercion-6) a few times to see that indeed `sample()` randomly generates numbers, so the integers in the vectors change in value each time. 

```{r numeric-coercion-5}
random_ii <- sample(1:15, size = 10) # create a vector of 10 random numbers between 1 and 15 
random_ii

class(random_ii)
typeof(random_ii)
mode(random_ii)
```

```{r numeric-coercion-6}
random_nd <- sample(1.4:15.8, size = 10) # create a vector of 10 random numbers between 1 and 15 
random_nd

class(random_nd)
typeof(random_nd)
mode(random_nd)
```

When converting a character object to numeric with `as.numeric()`, anything that cannot be parsed will turn into NA. R will let us know that NAs were coerced into the object with a warning. 

*📝 ⚠️⛔Note* Warnings are not equivalent to Errors in R. When there is an Error, typically this means the operation was not completed or the code not run. A Warning means the code was executed, but it may not be behaving as expected so you should check the output/result. Typically warnings are just that, warnings. But you should never ignore them completely. 

```{r numeric-coercion-7}
vector1 <- c("F", "1", "2", "4")
vector1

c2num <- as.numeric(vector1)
c2num
```


### 2.2 Built-in functions 

```{r num-round}
round(0.5) # rounds down as the nearest even digit is 0 

round(-1.5) # rounds up to -2 as it is the nearest even digit 

round(45.5) # rounds up 

round(46.5) # rounds down 

round(30.7) # rounds up 

round(90.2) # rounds down 
```

```{r num-abs}
abs(-47.6)

abs(367)
```

```{r num-summary}
num_obj <- c(1, 3, 4, 67.8, 90)

sum(num_obj)

min(num_obj)

max(num_obj)
```

### 2.3 Arithmetic operators 

There are 7 arithmetic operators in R. 

```{r arithmetic-operators}
# assign the values "7" and "3" to objects "xx" and "yy"
xx <- 7
yy <- 3 
```

```{r arithemtic-operators2}
# arithmetic operators 

## addition 
xx + yy 

## subtraction 
xx - yy

## division 
xx / yy  

## integer division/floor division
xx %/% yy

## exponent 
xx ^ yy

## modulus 
xx %% yy

```

As a statistical programming language, there is of course *tons* more than can be done in R with numeric data. We will see more of this as we explore different data structures next week. 

## You did it! 🎉

Well done for making it to the end of this notebook. If you have not done so yet, move to the Python numeric notebook next or the R character notebook. 

---
*Dr. Brittany Blankinship (2026)* 
